---
title: "Custom Models"
author: "Dimitris Rizopoulos"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Custom Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("GLMMadaptive")
```

# User-Defined Family Objects
In addition to the standard family objects in R, function `mixed_model()` also allows for 
user-specified family objects to be given in its `family` argument. The minimum 
requirements are that user-specified family object is of class `"family"` and is a list 
with the following components:

- `family`: a character string giving the name of the family.

- `link`: a character string giving the name of the link function.

- `linkfun`: a function that transforms the mean of the distribution of the outcome to the 
linear predictor scale.

- `linkinv`: a function that is the inverse of the `linkfun` defined above.

- `log_dens`: a function that returns the log of the probability density function or the 
log of the probability mass function of the distribution of the outcome given the random 
effects. This function should have exactly five arguments named, `y` the repeated 
measurements outcome, `eta` denoting the linear predictor (fixed + random effects), 
`mu_fun` the function to calculate the mean from the linear predictor (i.e., the `linkinv`
function), `phis` a vector of possible extra dispersion/scale parameters, and `eta_zi` the
linear predictor (fixed + random effects) for the logistic regression for the extra zeros; 
relevant in the cases of zero-inflated and two-part models. 
**Even if `phis` and `eta_zi` may not be required, it should always appear as arguments of `log_dens`**. 
Moreover, when the extra parameters `phis` are required, they should be appropriately 
transformed in order for the optimization of the log-likelihood to be unconstrained with
regard to these parameters. Finally, `log_dens` should be fully vectorized with respect to
`eta` and `eta_zi`. That is, for a numeric vector `y` and a numeric matrix `eta`, it 
should return a matrix with the log density evaluated at each column of  `eta`, and 
likewise for `eta_zi`.

For the internal calculations the derivative of `log_dens` with respect to the linear 
predictor `eta` and extra parameters `phis` (when present) is also required. If it is not
given, then it is approximated numerically using a central difference approximation. 
However, it greatly facilitates stability and speed if the user also specifies these 
functions. If the user specifies these functions, they should have the
names `score_eta_fun` and `score_phis_fun`, respectively, and both of them have three 
arguments named `y` the numeric vector of the repeated measurements outcome, `mu` the
numeric vector or matrix of the mean of the outcome and `phis` a numeric vector of the 
extra parameters.

## Negative Binomial Mixed Effects Model
We first illustrate the use of a custom-made family object in the case of a negative 
binomial repeated measurements outcome. We should note however that a family object
`negative.binomial()` is already defined in the package and users could invoke this 
instead - for more info see the relative online help page for `negative.binomial()`. We 
start by simulating some data:
```{r}
set.seed(101)
dd <- expand.grid(f1 = factor(1:3), f2 = LETTERS[1:2], g = 1:30, rep = 1:15,
                  KEEP.OUT.ATTRS = FALSE)
mu <- 5*(-4 + with(dd, as.integer(f1) + 4 * as.numeric(f2)))
dd$y <- rnbinom(nrow(dd), mu = mu, size = 0.5)
```

Next we define a new family object for negative binomial data
```{r}
my_negBinom <- function (link = "log") {
    stats <- make.link(link)
    log_dens <- function (y, eta, mu_fun, phis, eta_zi) {
        # the log density function
        phis <- exp(phis) # unconstrained optimization of 'phis'
        mu <- mu_fun(eta)
        log_mu_phis <- log(mu + phis)
        comp1 <- lgamma(y + phis) - lgamma(phis) - lgamma(y + 1)
        comp2 <- phis * log(phis) - phis * log_mu_phis
        comp3 <- y * log(mu) - y * log_mu_phis
        out <- comp1 + comp2 + comp3
        attr(out, "mu_y") <- mu
        out
    }
    structure(list(family = "user Neg Binom", link = stats$name, linkfun = stats$linkfun,
                   linkinv = stats$linkinv, log_dens = log_dens),
              class = "family")
}
```

And we fit the model
```{r}
system.time(
    fm1 <-  mixed_model(fixed = y ~ f1 * f2, random = ~ 1 | g, data = dd, 
                        family = my_negBinom(), n_phis = 1, 
                        initial_values = list("betas" = poisson()))
)

fm1
```

In the call to `mixed_model()` we specify the argument `n_phis` which lets the function
know how many extra parameters does the distribution of the outcome have. In addition, we 
also specify that better quality initial values for the fixed effects `betas` can be 
obtained by fitting a Poisson regression model to the repeated measurements outcome 
(ignoring the correlations) with the same design matrix as the fixed-effects design 
matrix. For the unique elements of the random effects covariance matrix and for the extra 
parameters `phis` the default initial values are used.

In the output, the reported phi parameters is on the log scale, because in the definition of
`log_dens` we specified that `phis <- exp(phis)`.

As mentioned above, the internal computations are facilitated if the user also specifies 
the derivatives of the log density with respect to the linear predictor `eta` and the
extra parameters `phis`. Hence, in the following we define a new negative binomial 
family object with these functions also defined:
```{r}
my_negBinom2 <- function () {
    stats <- make.link(link = "log")
    log_dens <- function (y, eta, mu_fun, phis, eta_zi) {
        # the log density function
        phis <- exp(phis)
        mu <- mu_fun(eta)
        log_mu_phis <- log(mu + phis)
        comp1 <- lgamma(y + phis) - lgamma(phis) - lgamma(y + 1)
        comp2 <- phis * log(phis) - phis * log_mu_phis
        comp3 <- y * log(mu) - y * log_mu_phis
        out <- comp1 + comp2 + comp3
        attr(out, "mu_y") <- mu
        out
    }
    score_eta_fun <- function (y, mu, phis, eta_zi) {
        # the derivative of the log density w.r.t. mu
        phis <- exp(phis)
        mu_phis <- mu + phis
        comp2 <- - phis / mu_phis
        comp3 <- y / mu - y / mu_phis
        # the derivative of mu w.r.t. eta (this depends on the chosen link function)
        mu.eta <- mu
        (comp2 + comp3) * mu.eta
    }
    score_phis_fun <- function (y, mu, phis, eta_zi) {
        # the derivative of the log density w.r.t. phis
        phis <- exp(phis)
        mu_phis <- mu + phis
        comp1 <- digamma(y + phis) - digamma(phis)
        comp2 <- log(phis) + 1 - log(mu_phis) - phis / mu_phis
        comp3 <- - y / mu_phis
        (comp1 + comp2 + comp3) * phis
    }
    structure(list(family = "user Neg Binom", link = stats$name, linkfun = stats$linkfun,
                   linkinv = stats$linkinv, log_dens = log_dens,
                   score_eta_fun = score_eta_fun, score_phis_fun = score_phis_fun),
              class = "family")
}
```

And we fit the model
```{r}
system.time(
    fm2 <-  mixed_model(fixed = y ~ f1 * f2, random = ~ 1 | g, data = dd, 
                        family = my_negBinom2(), n_phis = 1, 
                        initial_values = list("betas" = poisson()))
)

fm2
```

We obtain almost identical results, but in half of the computing time in this example.
Depending on the size of the data set, the gain in computing time when the derivatives 
for `eta` and `phis` are provided can be much more substantial.

## Student's t Mixed Effects Model
We further illustrate the use of custom-made family objects by fitting a linear mixed 
model but with Student's t distributed error terms. We start by simulating some data
```{r}
set.seed(1234)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 15 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ time, data = DF)

betas <- c(20.1, -0.5, 0.24, -0.05) # fixed effects coefficients
sigma <- 1.2 # scale parameter for the Student's t errors
D11 <- 1.5 # variance of random intercepts
D22 <- 1.2 # variance of random slopes

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)))
# linear predictor
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, ]))
# we simulate Student's t longitudinal data
DF$y <- eta_y + sigma * rt(n * K, df = 4)
```

Following we define the custom-made family object for the Student's t distribution:
```{r}
students.t <- function (df = stop("'df' must be specified"), link = "identity") {
    .df <- df
    env <- new.env(parent = .GlobalEnv)
    assign(".df", df, envir = env)
    stats <- make.link(link)
    log_dens <- function (y, eta, mu_fun, phis, eta_zi) {
        # the log density function
        sigma <- exp(phis)
        out <- dt(x = (y - eta) / sigma, df = .df, log = TRUE) - log(sigma)
        attr(out, "mu_y") <- eta
        out
    }
    score_eta_fun <- function (y, mu, phis, eta_zi) {
        # the derivative of the log density w.r.t. mu
        sigma2 <- exp(phis)^2
        y_mu <- y - mu
        (y_mu * (.df + 1) / (.df * sigma2)) / (1 + y_mu^2 / (.df * sigma2))
    }
    score_phis_fun <- function (y, mu, phis, eta_zi) {
        sigma <- exp(phis)
        y_mu2_df <- (y - mu)^2 / .df
        (.df + 1) * y_mu2_df * sigma^{-2} / (1 + y_mu2_df / sigma^2) - 1
    }
    environment(log_dens) <- environment(score_eta_fun) <- environment(score_phis_fun) <- env
    structure(list(family = "Student's-t", link = stats$name, linkfun = stats$linkfun,
                   linkinv = stats$linkinv, log_dens = log_dens, 
                   score_eta_fun = score_eta_fun, score_phis_fun = score_phis_fun),
              class = "family")
}
```

And we fit the model
```{r}
fm <-  mixed_model(y ~ sex * time, random = ~ time | id, data = DF, 
                   family = students.t(4), n_phis = 1, 
                   initial_values = list("betas" = gaussian()))

fm
```

As was the case also for the negative binomial model, the extra `phis` parameter has been
appropriately transformed to have an unconstrained optimization of the likelihood. Hence,
the estimated `sigma` parameter is `exp(fm$phis) =` `r unname(round(exp(fm$phis), 3))`. 

## Beta Mixed-Effects Model
As a final illustration we show how a beta mixed effects model can be fitted, i.e.,
applicable in case of bounded repeated measurements outcomes. We start again by
simulating some data:
```{r}
set.seed(1234)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 15 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ time, data = DF)

betas <- c(-2.2, -0.25, 0.24, -0.05) # fixed effects coefficients
phi <- 5 # precision parameter of the Beta distribution
D11 <- 0.9 # variance of random intercepts
D22 <- 0.3 # variance of random slopes

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)))
# linear predictor
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, ]))
# mean
mu <- plogis(eta_y)
# we simulate beta longitudinal data
DF$y <- rbeta(n * K, shape1 = mu * phi, shape2 = phi * (1 - mu))
# we transform to (0, 1)
DF$y <- (DF$y * (nrow(DF) - 1) + 0.5) / nrow(DF)
```

Next we define the custom-made family object as in the case of the Student's t 
distribution:
```{r}
beta.fam <- function () {
    stats <- make.link("logit")
    log_dens <- function (y, eta, mu_fun, phis, eta_zi) {
        # the log density function
        phi <- exp(phis)
        mu <- mu_fun(eta)
        mu_phi <- mu * phi
        comp1 <- lgamma(phi) - lgamma(mu_phi)
        comp2 <- (mu_phi - 1) * log(y) - lgamma(phi - mu_phi)
        comp3 <- (phi - mu_phi - 1) * log(1 - y)
        out <- comp1 + comp2 + comp3
        attr(out, "mu_y") <- mu
        out
    }
    score_eta_fun <- function (y, mu, phis, eta_zi) {
        # the derivative of the log density w.r.t. mu
        phi <- exp(phis)
        mu_phi <- mu * phi
        comp1 <- - digamma(mu_phi) * phi
        comp2 <- phi * (log(y) + digamma(phi - mu_phi))
        comp3 <- - phi * log(1 - y)
        # the derivative of mu w.r.t. eta (this depends on the chosen link function)
        mu.eta <- mu - mu * mu
        (comp1 + comp2 + comp3) * mu.eta
    }
    score_phis_fun <- function (y, mu, phis, eta_zi) {
        phi <- exp(phis)
        mu_phi <- mu * phi
        mu1 <- 1 - mu
        comp1 <- digamma(phi) - digamma(mu_phi) * mu
        comp2 <- mu * log(y) - digamma(phi - mu_phi) * mu1
        comp3 <- log(1 - y) * mu1
        (comp1 + comp2 + comp3) * phi
    }
    structure(list(family = "beta", link = stats$name, linkfun = stats$linkfun,
                   linkinv = stats$linkinv, log_dens = log_dens, 
                   score_eta_fun = score_eta_fun, score_phis_fun = score_phis_fun),
              class = "family")
}
```

And we fit the model
```{r}
gm <- mixed_model(y ~ sex * time, random = ~ time | id, data = DF,
                  family = beta.fam(), n_phis = 1)

gm
```

To depict the results of the model we create an effects plot, showing the longitudinal 
evolution of the average/median male and female. We start by constructing the data frame 
that contains the values we want to depict, and using it in the `effectPlotData()` 
function:
```{r}
nDF <- with(DF, expand.grid(time = seq(min(time), max(time), length = 25),
                            sex = levels(sex)))

plot_data <- effectPlotData(gm, nDF)
```

The figure is created with a call to `xyplot()` from the 
[**lattice**](https://cran.r-project.org/package=lattice) package:
```{r, fig.show = "hold"}
library("lattice")
xyplot(pred + low + upp ~ time | sex, data = plot_data,
       type = "l", lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2,
       xlab = "Follow-up Time", ylab = "Logit Scale")

expit <- function (x) exp(x) / (1 + exp(x))
xyplot(expit(pred) + expit(low) + expit(upp) ~ time | sex, data = plot_data,
       type = "l", lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2,
       xlab = "Follow-up Time", ylab = "Original Scale")
```
