---
title: "Methods for MixMod Objects"
author: "Dimitris Rizopoulos"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Methods for MixMod Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("GLMMadaptive")
```

# MixMod Objects
In this vignette we illustrate the use of a number of basic generic functions for models 
fitted by the `mixed_model()` function of package **GLMMadaptive**. 

We start by simulating some data for a binary longitudinal outcome:
```{r}
set.seed(1234)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 15 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ time, data = DF)

betas <- c(-2.13, -0.25, 0.24, -0.05) # fixed effects coefficients
D11 <- 0.48 # variance of random intercepts
D22 <- 0.1 # variance of random slopes

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)))
# linear predictor
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, ]))
# we simulate binary longitudinal data
DF$y <- rbinom(n * K, 1, plogis(eta_y))
```

We continue by fitting the mixed effects logistic regression for `y` assuming random
intercepts and random slopes for the random-effects part.
```{r}
fm <- mixed_model(fixed = y ~ sex * time, random = ~ time | id, data = DF,
                  family = binomial())
```

## Detailed Output, Confidence Intervals, Covariance Matrix of the MLEs and Robust Standard Errors
As in the majority of model-fitting functions in R, the `print()` and `summary()` methods
display a short and a detailed output of the fitted model, respectively. For `'MixMod'` 
objects we obtain

```{r}
fm
```

and

```{r}
summary(fm)
```

The output is rather self-explanatory. However, just note that the fixed-effects 
coefficients are on the linear predictor scale, and hence are the corresponding log-odds
for the intercept and log-odds ratios for the rest of the parameters. The `summary()` only
shows the estimated coefficients, standard errors and p-values, but no confidence 
intervals. These can be separately obtained using the `confint()` method, i.e.,
```{r}
exp(confint(fm))
```

By default the confidence intervals are produced for the fixed effects. Hence, taking the
exp we obtain the confidence intervals for the corresponding odds-ratios. In addition, by
default, the level of the confidence intervals is 95%. The following piece of code 
produces 90% confidence intervals for the variances of the random intercepts and slopes, 
and for their covariance:
```{r}
confint(fm, parm = "var-cov", level = 0.90)
```

The estimated variance-covariance matrix of the maximum likelihood estimates of all 
parameters is returned using the `vcov()` method, e.g.,
```{r}
vcov(fm)
```

The elements of this covariance matrix that correspond to the elements of the covariance 
matrix of the random effects (i.e., the elements `D_xx`) are on the log-Cholesky scale.

Robust standard errors using the sandwich estimator can be obtained by setting the logical
argument `sandwich` to 
`TRUE`, i.e.,
```{r}
vcov(fm, sandwich = TRUE)
```

The use of robust standard errors via the `sandwich` argument is also available in the 
`summary()`, `confint()`, `anova()`, `marginal_coefs()`, `effectPlotData()`, `predict()`, 
and `simulate()` methods.

## Fixed and Random Effects Estimates
To extract the estimated fixed effects coefficients from a fitted mixed model, we can use
the `fixef()` method. Similarly, the empirical Bayes estimates of the random effects are
extracted using the `ranef()` method, and finally the `coef()` method returns the 
subject-specific coefficients, i.e., the sum of the fixed and random effects coefficients:

```{r}
fixef(fm)
```

```{r}
head(ranef(fm))
```

```{r}
head(coef(fm))
```

## Marginalized Coefficients
The fixed effects estimates in mixed models with nonlinear link functions have an 
interpretation conditional on the random effects. However, often we wish to
obtain parameters with a marginal / population averaged interpretation, which leads many 
researchers to use generalized estimating equations, and dealing with potential issues
with missing data. Nonetheless, recently [Hedeker et al.](https://dx.doi.org/10.1111/biom.12707) 
have proposed a nice solution to this problem. Their approach is implemented in function
`marginal_coefs()`. For example, for model `fm` we obtain the marginalized coefficients
using:
```{r}
marginal_coefs(fm)
```

The function calculates the marginal log odds ratios in our case (because we have a binary
outcome) using a Monte Carlo procedure with number of samples determined by the `M` 
argument.

Standard errors for the marginalized coefficients are obtained by setting 
`std_errors = TRUE` in the call to `marginal_coefs()`, and require a double Monte Carlo
procedure for which argument `K` comes also into play. To speed up computations, the 
outer Monte Carlo procedure is performed in parallel using package **parallel** and 
number of cores specified in the `cores` argument (results not shown): 
```{r, eval = FALSE}
marginal_coefs(fm, std_errors = TRUE, cores = 5)
```


## Fitted Values and Residuals
The `fitted()` method extracts fitted values from the fitted mixed model. These are always
on the scale of the response variable. The `type` argument of `fitted()` specifies the 
type of fitted values computed. The default is `type = "mean_subject"` which corresponds 
to the fitted values calculated using only the fixed-effects part of the linear predictor;
hence, for the subject who has random effects values equal to 0, i.e., the "mean subject":
```{r}
head(fitted(fm))
```

Setting `type = "subject_specific"` will calculate the fitted values using both the fixed
and random effects parts, where for the latter the empirical Bayes estimates of the random
effects are used:
```{r}
head(fitted(fm, type = "subject_specific"))
```

Finally, setting `type = "marginal"` will calculate the fitted values based on 
the multiplication of the fixed-effects design matrix with the marginalized coefficients 
described above (due to the required computing time, these fitted values are not displayed):
```{r, eval = FALSE}
head(fitted(fm, type = "marginal"))
```

The `residuals()` method simply calculates the residuals by subtracting the fitted values
from the observed repeated measurements outcome. Hence, this method also has a `type` 
argument with exactly the same options as the `fitted()` method.

## Effect Plots
To display the estimated longitudinal evolutions of the binary outcome we can use an 
effect plot. This is simply predictions from the models with the corresponding 95% 
pointwise confidence intervals. 

As a first step we create a data frame the provides the setting based on which the plot
is to be produced; function `expand.grid()` is helpful in this regard:
```{r}
nDF <- with(DF, expand.grid(time = seq(min(time), max(time), length.out = 15),
                            sex = levels(sex)))
```

Next we use the `effectPlotData()` function that does the heavy lifting, i.e., calculates
the predictions and confidence intervals from a fitted mixed model for the data frame 
provided above, i.e.,
```{r}
plot_data <- effectPlotData(fm, nDF)
```

Then we can produce the plot using for example the `xyplot()` function from package 
**lattice**, e.g.,
```{r, fig.show = "hold", fig.align = "center", fig.width = 8.5, fig.height = 7.5}
library("lattice")
xyplot(pred + low + upp ~ time | sex, data = plot_data,
       type = "l", lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2,
       xlab = "Follow-up time", ylab = "log odds")

expit <- function (x) exp(x) / (1 + exp(x))
xyplot(expit(pred) + expit(low) + expit(upp) ~ time | sex, data = plot_data,
       type = "l", lty = c(1, 2, 2), col = c(2, 1, 1), lwd = 2,
       xlab = "Follow-up time", ylab = "Subject-Specific Probabilities")
```

The `effectPlotData()` function also allows to compute marginal predictions using the
marginalized coefficients described above. This is achieved by setting `marginal = TRUE`
in  the respective call (results not shown):
```{r, fig.align = "center", fig.width = 8.5, fig.height = 7.5, eval = FALSE}
plot_data_m <- effectPlotData(fm, nDF, marginal = TRUE)

# we put the two groups in the same panel
my.panel.bands <- function(x, y, upper, lower, fill, col, subscripts, ..., font, 
                           fontface) {
    upper <- upper[subscripts]
    lower <- lower[subscripts]
    panel.polygon(c(x, rev(x)), c(upper, rev(lower)), col = fill, border = FALSE, ...)
}

xyplot(expit(pred) ~ time, group = sex, data = plot_data_m, 
       upper = expit(plot_data_m$upp), low = expit(plot_data_m$low), 
       type = "l", col = c("blue", "red"), 
       fill = c("#0000FF80", "#FF000080"),
       panel = function (x, y, ...) {
           panel.superpose(x, y, panel.groups = my.panel.bands, ...)
           panel.xyplot(x, y, lwd = 2,  ...)
}, xlab = "Follow-up time", ylab = "Marginal Probabilities")
```

## Effect Plots using the effects package
In addition to using the `effectPlotData()` function, the same type of effect plots can
be produced by the [**effects**](https://cran.r-project.org/package=effects) package. For
example, based on the `fm` model we produce the effect plot for `time` and for the two 
groups using the call to `predictorEffect()` and its `plot()` method:
```{r, fig.align = "center", fig.width = 8.5, fig.height = 7.5, message = FALSE}
library("effects")
plot(predictorEffect("time", fm), type = "link")
```

The `type` argument controls the scale of the y-axis, namely, `type = "link"` corresponds 
to the log-odds scale. If we would like to obtain the figure in the probability scale, we
should set `type = "response"`, i.e.,
```{r, fig.align = "center", fig.width = 8.5, fig.height = 7.5, message = FALSE}
plot(predictorEffect("time", fm), type = "response")
```

For the additional functionality provided by the [**effects**](https://cran.r-project.org/package=effects) package, check the vignette: `vignette("predictor-effects-gallery", package = "effects")`.

**Note:** Effects plots via the **effects** package are currently only supported for the `binomial()` and `poisson()` families.

## Comparing Two Models
The `anova()` method can be used to compare two fitted mixed models using a likelihood 
ratio test. For example, we test if we can test the null hypothesis that the covariance
between the random intercepts and slopes is equal to zero using
```{r}
gm <- mixed_model(fixed = y ~ sex * time, random = ~ time || id, data = DF,
                  family = binomial())

anova(gm, fm)
```


## Predictions
Using the `predict()` method we can calculate predictions for new subjects. As an example,
we treat subject 1 from the `DF` dataset as a new patient (in the code below we change his
`id` variable):
```{r}
pred_DF <- DF[DF$id == 1, ][1:4, ]
pred_DF$id <- paste0("N", as.character(pred_DF$id))
pred_DF
```

We start by computing predictions based only on the fixed-effects part of the model; 
because of the nonlinear link function, these predictions are of subjects with random 
effects value equal to zero, which is not to the average predictions:
```{r}
predict(fm, newdata = pred_DF, type_pred = "response",
        type = "mean_subject", se.fit = TRUE)
```

Population averaged predictions can be obtained by first calculating the marginalized 
coefficients (using `marginal_coefs()`) and multiplying them with the fixed effects 
design matrix; this is achieved using the option `type = "marginal"` (due to the required 
computing time, predictions not shown):
```{r, eval = FALSE}
predict(fm, newdata = pred_DF, type_pred = "response",
        type = "marginal", se.fit = FALSE)
```

Finally, we calculate subject-specific predictions; the standard errors are calculated 
with a Monte Carlo scheme (for details check the online help file):
```{r}
predict(fm, newdata = pred_DF, type_pred = "response",
        type = "subject_specific", se.fit = TRUE)
```

Suppose now that we want predictions at time points at which no responses `y` have
been recorded, e.g.,
```{r}
future_Times <- pred_DF[1:3, c("id", "time", "sex")]
future_Times$time <- c(3, 4, 10)
future_Times
```

Predictions at these time points can be calculated by provide this data frame in the 
`newdata2` argument of `predict()`:
```{r}
predict(fm, newdata = pred_DF, newdata2 = future_Times, type_pred = "response",
        type = "subject_specific", se.fit = TRUE)
```

## Simulate
The `simulate()` method can be used to simulate response outcome data from a fitted mixed 
model. For example, we simulate two realization of our dichotomous outcome:
```{r}
head(simulate(fm, nsim = 2, seed = 123), 10)
```

By setting `acount_MLEs_var = TRUE` in the call to `simulate()` we also account for the 
variability in the maximum likelihood estimates in the simulation of new responses. This
is achieved by simulating each time a new response vector using a realization for the 
parameters from a multivariate normal distribution with mean the MLEs and covariance 
matrix the covariance matrix of the MLEs:
```{r}
head(simulate(fm, nsim = 2, seed = 123, acount_MLEs_var = TRUE), 10)
```
