---
title: "Zero-Inflated and Two-Part Mixed Effects Models"
author: "Dimitris Rizopoulos"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Zero-Inflated and Two-Part Mixed Effects Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("GLMMadaptive")
```

# Mixed Models with an Extra Zero Part
Function `mixed_model()` of **GLMMadaptive** can also be used to fit zero-inflated and 
two-part mixed effects models. For both types of models, a suitable `family` object needs 
to be specified as outlined in the vignette 
[Custom Models](http://www.drizopoulos.com/vignettes/Custom_Models.html), and also 
arguments `zi_fixed` and `zi_random` of `mixed_model()` come into play. In these 
arguments, the user can specify the fixed and random effects `formulas` of the logistic 
regression for the zero-part of the distribution of the outcome. We should note that the 
user has the option to leave `zi_random` set to `NULL`, in which case the for the zero-part
we have a logistic regression with only fixed effect and no random effects.

In addition, in the specification of the `family` object and in order to better facilitate 
the internal computations, the user may specify the function `score_eta_zi_fun` that 
calculates the derivative of the log probability density function or the log probability
mass function with respect to `eta_zi` that denotes the linear predictor of the logistic 
regression for the zero-part. Here we provide three examples to illustrate how such models
can be fitted.

## Zero-Inflated Poisson Mixed Effects Model
We start our illustrations by showing how we can fit a zero-inflated Poisson mixed effects
model. The specification of the required `family` object is already available in the 
package as the object returned by `zi.poisson()`. *Currently, only the log link is allowed.* 
First, we simulate longitudinal data from a zero-inflated negative binomial distribution:
```{r}
set.seed(1234)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 5 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects non-zero part
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ 1, data = DF)
# design matrices for the fixed and random effects zero part
X_zi <- model.matrix(~ sex, data = DF)
Z_zi <- model.matrix(~ 1, data = DF)

betas <- c(1.5, 0.05, 0.05, -0.03) # fixed effects coefficients non-zero part
shape <- 2
gammas <- c(-1.5, 0.5) # fixed effects coefficients zero part
D11 <- 0.5 # variance of random intercepts non-zero part
D22 <- 0.4 # variance of random intercepts zero part

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)))
# linear predictor non-zero part
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, 1, drop = FALSE]))
# linear predictor zero part
eta_zi <- as.vector(X_zi %*% gammas + rowSums(Z_zi * b[DF$id, 2, drop = FALSE]))
# we simulate negative binomial longitudinal data
DF$y <- rnbinom(n * K, size = shape, mu = exp(eta_y))
# we set the extra zeros
DF$y[as.logical(rbinom(n * K, size = 1, prob = plogis(eta_zi)))] <- 0
```

A zero-inflated Poisson mixed model with only fixed effects in the zero part is fitted 
with the following call to `mixed_model()`:
```{r}
fm1 <- mixed_model(y ~ sex * time, random = ~ 1 | id, data = DF,
                   family = zi.poisson(), zi_fixed = ~ sex)

fm1
```

As noted above, only the log link is currently available for the non-zero part and the 
logit link for the zero-part. Hence, the estimated fixed effects for the two parts are 
interpreted accordingly. We extend `fm1` by allowing also for random intercept in the 
zero-part. We should note that by default the random intercept from the non-zero part is 
correlated with the random intercept from the zero-part:
```{r}
fm2 <- update(fm1, zi_random = ~ 1 | id)

fm2
```

We test if we need the extra random effect using a likelihood ratio test:
```{r}
anova(fm1, fm2)
```

The results suggest that the extra random effect improves the fit of the model.

## Zero-Inflated Negative Binomial Mixed Effects Model
We continue with the same data, but we now take into account the potential over-dispersion 
in the data using a zero-inflated negative binomial model. To fit this mixed model we use 
almost identical syntax to what just did above - the only difference that now we specify
as family the `zi.negative.binomial()` object:
```{r}
gm1 <- mixed_model(y ~ sex * time, random = ~ 1 | id, data = DF,
                   family = zi.negative.binomial(), zi_fixed = ~ sex)

gm1
```

Similarly to `fm1` in `gm1` we specified only fixed effects for the logistic regression 
for the zero part. We now compare this model with the zero-inflated Poisson model that 
allowed for a random intercept in the zero-part. The comparison can be done with the
`anova()` method; because the two models are not nested we set `test = FALSE` in the call
to `anova()`, i.e.:
```{r}
anova(gm1, fm2, test = FALSE)
```

We observe that accounting for the over-dispersion seems to better improve the fit than
including the random intercepts term in the zero-part.

## Two-Part Mixed Effects Model for Semi-Continuous Data
To further illustrate the flexibility provided by **GLMMadaptive** in allowing users to
specify their own family objects with a specific log density function, we consider the 
setting of multivariate semi-continuous data. That is, continuous data with excess zeros.
In the literature the class of two-part / hurdle mixed models has been proposed to analyze
such data. These models specify a logistic regression for the dichotomous indicator that 
the outcome is zero or not, and a standard linear mixed model for the logarithmic 
transformation of the non-zero responses.

We start again by simulating some longitudinal data from this model:
```{r}
set.seed(1234)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 5 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects non-zero part
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ time, data = DF)
# design matrices for the fixed and random effects zero part
X_zi <- model.matrix(~ sex, data = DF)
Z_zi <- model.matrix(~ 1, data = DF)

betas <- c(-2.13, -0.25, 0.24, -0.05) # fixed effects coefficients non-zero part
sigma <- 0.5 # standard deviation error terms non-zero part
gammas <- c(-1.5, 0.5) # fixed effects coefficients zero part
D11 <- 0.5 # variance of random intercepts non-zero part
D22 <- 0.1 # variance of random slopes non-zero part
D33 <- 0.4 # variance of random intercepts zero part

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)), rnorm(n, sd = sqrt(D33)))
# linear predictor non-zero part
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, 1:2, drop = FALSE]))
# linear predictor zero part
eta_zi <- as.vector(X_zi %*% gammas + rowSums(Z_zi * b[DF$id, 3, drop = FALSE]))
# we simulate log-normal longitudinal data
DF$y <- exp(rnorm(n * K, mean = eta_y, sd = sigma))
# we set the zeros from the logistic regression
DF$y[as.logical(rbinom(n * K, size = 1, prob = plogis(eta_zi)))] <- 0
```

We define the `family` object: The minimal requirement is to specify the `log_dens` 
component and the link function; however, as also explained in the [Custom Models](http://www.drizopoulos.com/vignettes/Custom_Models.html) vignette, the internal
calculations will be faster and more stable if the user also specifies the score vector 
for the linear predictor of the non-zero part (function `score_eta_fun()`), the derivative
of the log density with respect to `phis` (function `score_phis_fun()`), and because we 
have a model with a zero-part, also the derivative of the log density with respect to the
linear predictor of the zero part (function `score_eta_zi_fun()`). Finally, for being 
able to simulate from the model using the `simulate()` method, the function `simulate()`
within the family object can also be specified. Hence, the family object for the two-part 
model is defined as:
```{r}
two_part.lognormal <- function () {
    stats <- make.link("identity")
    log_dens <- function (y, eta, mu_fun, phis, eta_zi) {
        sigma <- exp(phis)
        # binary indicator for y > 0
        ind <- y > 0
        # non-zero part
        eta <- as.matrix(eta)
        eta_zi <- as.matrix(eta_zi)
        out <- eta
        out[ind, ] <- plogis(eta_zi[ind, ], lower.tail = FALSE, log.p = TRUE) + 
            dnorm(x = log(y[ind]), mean = eta[ind, ], sd = sigma, log = TRUE)
        # zero part
        out[!ind, ] <- plogis(eta_zi[!ind, ], log.p = TRUE)
        attr(out, "mu_y") <- eta
        out
    }
    score_eta_fun <- function (y, mu, phis, eta_zi) {
        sigma <- exp(phis)
        # binary indicator for y > 0
        ind <- y > 0
        # non-zero part
        eta <- as.matrix(mu)
        out <- eta
        out[!ind, ] <- 0
        out[ind, ] <- (log(y[ind]) - eta[ind, ]) / sigma^2
        out
    }
    score_eta_zi_fun <- function (y, mu, phis, eta_zi) {
        ind <- y > 0
        probs <- plogis(as.matrix(eta_zi))
        out <- 1 - probs
        out[ind, ] <- - probs[ind, ]
        out
    }
    score_phis_fun <- function (y, mu, phis, eta_zi) {
        sigma <- exp(phis)
        # binary indicator for y > 0
        ind <- y > 0
        # non-zero part
        eta <- as.matrix(mu)
        out <- eta
        out[!ind, ] <- 0
        out[ind, ] <- - 1 + (log(y[ind]) - eta[ind, ])^2 / sigma^2
        out
    }
    simulate <- function (n, mu, phis, eta_zi) {
        y <- rnorm(n = n, mean = mu, sd = exp(phis))
        y[as.logical(rbinom(n, 1, plogis(eta_zi)))] <- 0
        y
    }
    structure(list(family = "two-part log-normal", link = stats$name, 
                   linkfun = stats$linkfun, linkinv = stats$linkinv, log_dens = log_dens,
                   score_eta_fun = score_eta_fun, score_eta_zi_fun = score_eta_zi_fun,
                   score_phis_fun = score_phis_fun, simulate = simulate),
              class = "family")
}
```

Then to fit the model we provide the user-defined family object in the `family` argument 
of `mixed_model()` specifying also that we have one dispersion parameter in the family 
(i.e., `n_phis = 1`), and that in the zero-part we only include fixed effects: 
```{r}
km1 <- mixed_model(y ~ sex * time, random = ~ 1 | id, data = DF, 
                  family = two_part.lognormal(), n_phis = 1,
                  zi_fixed = ~ sex)

km1
```

The estimated standard deviation for the error terms is `exp(phis) = ` `r round(exp(km1$phis), 1)`
We extend the model by allowing for a random intercept in the zero-part, but using the
`||` symbol in the `random` argument we specify the covariance matrix of the random 
effects is diagonal; hence, that the two random intercepts terms are uncorrelated:
```{r}
km2 <- update(km1, random = ~ 1 || id, zi_random = ~ 1 | id)

km2
```

Finally, we show how the `simulate()` method can be used to perform a replication / 
posterior predictive check. In particular, in the following code we compare the empirical
distribution function estimated in the observed data with estimates of the empirical
distribution function obtained from simulated/replicated data from the model. In the call
to `simulate()` below we also specify to account for the variability in the maximum likelihood 
estimates by setting `acount_MLEs_var = TRUE`:
```{r}
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.1, 0.5, 0), cex.axis = 0.7, cex.lab = 0.8)
y <- DF$y
y[y > 0] <- log(y[y > 0])
x_vals <- seq(min(y), max(y), length.out = 500)
out <- simulate(km2, nsim = 30, acount_MLEs_var = TRUE)
rep_y <- apply(out, 2, function (x, x_vals) ecdf(x)(x_vals), x_vals = x_vals)
matplot(x_vals, rep_y, type = "l", lty = 1, col = "lightgrey", 
        xlab = "Response Variable", ylab = "Empirical CDF")
lines(x_vals, ecdf(y)(x_vals))
legend("bottomright", c("replicated data", "observed data"), lty = 1, 
       col = c("lightgrey", "black"), bty = "n", cex = 0.8)
```

