---
title: "Zero-Inflated and Two-Part Mixed Effects Models"
author: "Dimitris Rizopoulos"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Zero-Inflated and Two-Part Mixed Effects Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("GLMMadaptive")
```

# Mixed Models with an Extra Zero Part
Function `mixed_model()` of **GLMMadaptive** can also be used to fit zero-inflated and 
two-part mixed effects models. For both types of models, a suitable `family` object needs 
to be specified as outlined in the vignette 
[Custom Models](http://www.drizopoulos.com/vignettes/Custom_Models.html), and also 
arguments `zi_fixed` and `zi_random` of `mixed_model()` come into play. In these 
arguments, the user can specify the fixed and random effects `formulas` of the logistic 
regression for the zero-part of the distribution of the outcome. We should note that the 
user has the option to leave `zi_random` set to `NULL`, in which case for the zero-part
we have a logistic regression with only fixed effects and no random effects.

In addition, in the specification of the `family` object, and in order to better facilitate 
the internal computations, the user may specify the function `score_eta_zi_fun` that 
calculates the derivative of the log probability density function or the log probability
mass function with respect to `eta_zi` that denotes the linear predictor of the logistic 
regression for the zero part. Here we provide three examples to illustrate how such models
can be fitted.

## Zero-Inflated Poisson Mixed Effects Model
We start our illustrations by showing how we can fit a zero-inflated Poisson mixed effects
model. The specification of the required `family` object is already available in the 
package as the object returned by `zi.poisson()`. *Currently, only the log link is allowed.* 
First, we simulate longitudinal data from a zero-inflated negative binomial distribution:
```{r}
set.seed(1234)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 5 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects non-zero part
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ 1, data = DF)
# design matrices for the fixed and random effects zero part
X_zi <- model.matrix(~ sex, data = DF)
Z_zi <- model.matrix(~ 1, data = DF)

betas <- c(1.5, 0.05, 0.05, -0.03) # fixed effects coefficients non-zero part
shape <- 2 # shape/size parameter of the negative binomial distribution
gammas <- c(-1.5, 0.5) # fixed effects coefficients zero part
D11 <- 0.5 # variance of random intercepts non-zero part
D22 <- 0.4 # variance of random intercepts zero part

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)))
# linear predictor non-zero part
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, 1, drop = FALSE]))
# linear predictor zero part
eta_zi <- as.vector(X_zi %*% gammas + rowSums(Z_zi * b[DF$id, 2, drop = FALSE]))
# we simulate negative binomial longitudinal data
DF$y <- rnbinom(n * K, size = shape, mu = exp(eta_y))
# we set the extra zeros
DF$y[as.logical(rbinom(n * K, size = 1, prob = plogis(eta_zi)))] <- 0
```

A zero-inflated Poisson mixed model with only fixed effects in the zero part is fitted 
with the following call to `mixed_model()`:
```{r}
fm1 <- mixed_model(y ~ sex * time, random = ~ 1 | id, data = DF,
                   family = zi.poisson(), zi_fixed = ~ sex)

fm1
```

As noted above, only the log link is currently available for the non-zero part and the 
logit link for the zero part. Hence, the estimated fixed effects for the two parts are 
interpreted accordingly. We extend `fm1` by allowing also for random intercepts in the 
zero part. We should note that by default the random intercept of the non-zero part is 
correlated with the random intercept from the zero part:
```{r}
fm2 <- update(fm1, zi_random = ~ 1 | id)

fm2
```

We test if we need the extra random effect using a likelihood ratio test:
```{r}
anova(fm1, fm2)
```

The results suggest that the extra random effect improves the fit of the model.

## Zero-Inflated Negative Binomial Mixed Effects Model
We continue with the same data, but we now take into account the potential over-dispersion 
in the data using a zero-inflated negative binomial model. To fit this mixed model we use 
an almost identical syntax to what we just did above - the only difference is that we now 
specify as family the `zi.negative.binomial()` object:
```{r}
gm1 <- mixed_model(y ~ sex * time, random = ~ 1 | id, data = DF,
                   family = zi.negative.binomial(), zi_fixed = ~ sex)

gm1
```

Similarly to `fm1`, in `gm1` we specified only fixed effects for the logistic regression 
for the zero part. We now compare this model with the zero-inflated Poisson model that 
allowed for a random intercept in the zero part. The comparison can be done with the
`anova()` method; because the two models are not nested, we set `test = FALSE` in the call
to `anova()`, i.e.:
```{r}
anova(gm1, fm2, test = FALSE)
```

We observe that accounting for the over-dispersion seems to better improve the fit than
including the random intercepts term in the zero part.

## Two-Part Mixed Effects Model for Semi-Continuous Data
To further illustrate the flexibility provided by **GLMMadaptive** in allowing users to
specify their own family objects with a specific log density function, we consider the 
setting of multivariate semi-continuous data. That is, continuous data with excess zeros.
In the literature the class of two-part / hurdle mixed models has been proposed to analyze
such data. These models specify a logistic regression for the dichotomous indicator that 
the outcome is zero or not, and a standard linear mixed model for the logarithmic 
transformation of the non-zero responses.

We start again by simulating some longitudinal data from this model:
```{r}
set.seed(1234)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 5 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects non-zero part
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ time, data = DF)
# design matrices for the fixed and random effects zero part
X_zi <- model.matrix(~ sex, data = DF)
Z_zi <- model.matrix(~ 1, data = DF)

betas <- c(-2.13, -0.25, 0.24, -0.05) # fixed effects coefficients non-zero part
sigma <- 0.5 # standard deviation error terms non-zero part
gammas <- c(-1.5, 0.5) # fixed effects coefficients zero part
D11 <- 0.5 # variance of random intercepts non-zero part
D22 <- 0.1 # variance of random slopes non-zero part
D33 <- 0.4 # variance of random intercepts zero part

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)), rnorm(n, sd = sqrt(D33)))
# linear predictor non-zero part
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, 1:2, drop = FALSE]))
# linear predictor zero part
eta_zi <- as.vector(X_zi %*% gammas + rowSums(Z_zi * b[DF$id, 3, drop = FALSE]))
# we simulate log-normal longitudinal data
DF$y <- exp(rnorm(n * K, mean = eta_y, sd = sigma))
# we set the zeros from the logistic regression
DF$y[as.logical(rbinom(n * K, size = 1, prob = plogis(eta_zi)))] <- 0
```

To fit the two-part mixed model for log-normal data we can use the already build-in 
`hurdle.lognormal()` family object. However, just as an illustration, and to show that 
users can define their own family objects to be used in `mixed_model()`, we explain how
exactly `hurdle.lognormal()` is specified. To define the family object: The minimal 
requirement is to specify the `log_dens` 
component and the link function; however, as also explained in the [Custom Models](http://www.drizopoulos.com/vignettes/Custom_Models.html) vignette, the internal
calculations will be faster and more stable if the user also specifies the score vector 
for the linear predictor of the non-zero part (function `score_eta_fun()`), the derivative
of the log density with respect to `phis` (function `score_phis_fun()`), and because we 
have a model with a zero part, also the derivative of the log density with respect to the
linear predictor of the zero part (function `score_eta_zi_fun()`). Finally, for being 
able to simulate from the model using the `simulate()` method, the function `simulate()`
within the family object can also be specified. Hence, the family object for the two-part 
model is defined as:
```{r}
hurdle.lognormal <- function () {
    stats <- make.link("identity")
    log_dens <- function (y, eta, mu_fun, phis, eta_zi) {
        sigma <- exp(phis)
        # binary indicator for y > 0
        ind <- y > 0
        # non-zero part
        eta <- as.matrix(eta)
        eta_zi <- as.matrix(eta_zi)
        out <- eta
        out[ind, ] <- plogis(eta_zi[ind, ], lower.tail = FALSE, log.p = TRUE) + 
            dnorm(x = log(y[ind]), mean = eta[ind, ], sd = sigma, log = TRUE)
        # zero part
        out[!ind, ] <- plogis(eta_zi[!ind, ], log.p = TRUE)
        attr(out, "mu_y") <- eta
        out
    }
    score_eta_fun <- function (y, mu, phis, eta_zi) {
        sigma <- exp(phis)
        # binary indicator for y > 0
        ind <- y > 0
        # non-zero part
        eta <- as.matrix(mu)
        out <- eta
        out[!ind, ] <- 0
        out[ind, ] <- (log(y[ind]) - eta[ind, ]) / sigma^2
        out
    }
    score_eta_zi_fun <- function (y, mu, phis, eta_zi) {
        ind <- y > 0
        probs <- plogis(as.matrix(eta_zi))
        out <- 1 - probs
        out[ind, ] <- - probs[ind, ]
        out
    }
    score_phis_fun <- function (y, mu, phis, eta_zi) {
        sigma <- exp(phis)
        # binary indicator for y > 0
        ind <- y > 0
        # non-zero part
        eta <- as.matrix(mu)
        out <- eta
        out[!ind, ] <- 0
        out[ind, ] <- - 1 + (log(y[ind]) - eta[ind, ])^2 / sigma^2
        out
    }
    simulate <- function (n, mu, phis, eta_zi) {
        y <- rnorm(n = n, mean = mu, sd = exp(phis))
        y[as.logical(rbinom(n, 1, plogis(eta_zi)))] <- 0
        y
    }
    structure(list(family = "two-part log-normal", link = stats$name, 
                   linkfun = stats$linkfun, linkinv = stats$linkinv, log_dens = log_dens,
                   score_eta_fun = score_eta_fun, score_eta_zi_fun = score_eta_zi_fun,
                   score_phis_fun = score_phis_fun, simulate = simulate),
              class = "family")
}
```

Then to fit the model, we provide the user-defined family object in the `family` argument 
of `mixed_model()`, specifying also that we have one dispersion parameter in the family 
(i.e., `n_phis = 1`), and that in the zero part we only include fixed effects: 
```{r}
km1 <- mixed_model(y ~ sex * time, random = ~ 1 | id, data = DF, 
                  family = hurdle.lognormal(), n_phis = 1,
                  zi_fixed = ~ sex)

km1
```

The estimated standard deviation for the error terms is `exp(phis) = ` `r round(exp(km1$phis), 1)`
We extend the model by allowing for a random intercept in the zero-part, but using the
`||` symbol in the `random` argument we specify the covariance matrix of the random 
effects is diagonal; hence, that the two random intercepts terms are uncorrelated:
```{r}
km2 <- update(km1, random = ~ 1 || id, zi_random = ~ 1 | id)

km2
```

Finally, we show how the `simulate()` method can be used to perform a replication / 
posterior predictive check. In particular, in the following code we compare the empirical
distribution function estimated in the observed data with estimates of the empirical
distribution function obtained from simulated/replicated data from the model. In the call
to `simulate()` below we also specify to account for the variability in the maximum 
likelihood estimates by setting `acount_MLEs_var = TRUE`:
```{r}
par(mar = c(2.5, 2.5, 0, 0), mgp = c(1.1, 0.5, 0), cex.axis = 0.7, cex.lab = 0.8)
y <- DF$y
y[y > 0] <- log(y[y > 0])
x_vals <- seq(min(y), max(y), length.out = 500)
out <- simulate(km2, nsim = 30, acount_MLEs_var = TRUE)
ind <- out > sqrt(.Machine$double.eps)
out[ind] <- log(out[ind])
rep_y <- apply(out, 2, function (x, x_vals) ecdf(x)(x_vals), x_vals = x_vals)
matplot(x_vals, rep_y, type = "l", lty = 1, col = "lightgrey", 
        xlab = "Response Variable", ylab = "Empirical CDF")
lines(x_vals, ecdf(y)(x_vals))
legend("bottomright", c("log replicated data", "log observed data"), lty = 1, 
       col = c("lightgrey", "black"), bty = "n", cex = 0.8)
```

## Two-Part/Hurdle Poisson Mixed Effects Model
An alternative modeling framework to account for high percentages of 0 in count data is 
hurdle models. These models are similar to the two-part model for semi-continuous data 
presented above. The difference is that for the positive part we have positive counts 
instead of a positive continuous outcome. For the positive counts, a truncated at zero 
Poisson or negative binomial distribution is typically used. Both hurdle Poisson and hurdle 
negative binomial mixed models can be fitted by `mixed_model()` using the family objects
`hurdle.poisson()` and `hurdle.negative.binomial`, respectively.

To illustrate how these models are fitted, we simulate some longitudinal data from a 
hurdle negative binomial model using the code:
```{r}
set.seed(123)
n <- 100 # number of subjects
K <- 8 # number of measurements per subject
t_max <- 5 # maximum follow-up time

# we constuct a data frame with the design: 
# everyone has a baseline measurment, and then measurements at random follow-up times
DF <- data.frame(id = rep(seq_len(n), each = K),
                 time = c(replicate(n, c(0, sort(runif(K - 1, 0, t_max))))),
                 sex = rep(gl(2, n/2, labels = c("male", "female")), each = K))

# design matrices for the fixed and random effects non-zero part
X <- model.matrix(~ sex * time, data = DF)
Z <- model.matrix(~ time, data = DF)
# design matrices for the fixed and random effects zero part
X_zi <- model.matrix(~ sex, data = DF)
Z_zi <- model.matrix(~ 1, data = DF)

betas <- c(1.5, 0.05, 0.05, -0.03) # fixed effects coefficients non-zero part
shape <- 2 # shape/size parameter of the negative binomial distribution
gammas <- c(-1.5, 0.5) # fixed effects coefficients zero part
D11 <- 0.5 # variance of random intercepts non-zero part
D22 <- 0.1 # variance of random slopes non-zero part
D33 <- 0.4 # variance of random intercepts zero part

# we simulate random effects
b <- cbind(rnorm(n, sd = sqrt(D11)), rnorm(n, sd = sqrt(D22)), rnorm(n, sd = sqrt(D33)))
# linear predictor non-zero part
eta_y <- as.vector(X %*% betas + rowSums(Z * b[DF$id, 1:2, drop = FALSE]))
# linear predictor zero part
eta_zi <- as.vector(X_zi %*% gammas + rowSums(Z_zi * b[DF$id, 3, drop = FALSE]))
# we simulate truncated at zero negative binomial longitudinal data
lower <- pnbinom(0, mu = exp(eta_y), size = shape)
u <- runif(n * K, min = lower, max = 1)
DF$y <- qnbinom(u, mu = exp(eta_y), size = shape)
# we set the zeros from the logistic regression
DF$y[as.logical(rbinom(n * K, size = 1, prob = plogis(eta_zi)))] <- 0
```

The following code fits a hurdle Poisson mixed effects model. In the fixed-effects part 
for the positive counts we include the main effects of `sex` and `time` and their 
interaction, and in the random-effects for the positive counts random intercepts and random
slopes. For the zero-part we only include a fixed-effects part, using argument `zi_fixed`, 
with `sex` as the predictor: 
```{r}
dm1 <- mixed_model(y ~ sex * time, random = ~ time | id, data = DF, 
                  family = hurdle.poisson(), zi_fixed = ~ sex)

dm1
```

Currently, only the log-link is implemented for the `hurdle.poisson()` family. The output
is similar as in the case of the zero-inflated models. However, it has been noted that 
the fixed-effects coefficients for the positive counts part relate to the mean $\mu$ of
Poisson distribution that includes the zeros, not the mean for those who experience the 
event. The mean for those who experience more than one events is $\mu / (1 - e^{-\mu})$. 
Hence, a $\beta = 0.1$ cannot be interpreted as reflecting a $e^\beta =$ 10% increase in 
the mean of the subjects who experience the events.

We extend model `dm1` by also including a random intercept for the zero-part. As we have 
done above, this is achieved by specifying the `zi_random` argument, i.e.:
```{r}
dm2 <- update(dm1, zi_random = ~ 1 | id)

dm2
```

The likelihood ratio test between the two models is computed with the `anova()` method:
```{r}
anova(dm1, dm2)
```

The result indicates that the extra random effect for the zero-part improves the fit of
the model.

## Two-Part/Hurdle Negative Binomial Mixed Effects Model
We continue our illustration of hurdle models by fitting the hurdle negative binomial 
mixed model. The only change from the previous syntax we used is the name of the family 
object, namely, we use now `hurdle.negative.binomial()`. The code below fits exactly the
same model as model `dm1` above, but using now the negative binomial distribution for the
positive counts:
```{r}
hm1 <- mixed_model(y ~ sex * time, random = ~ time | id, data = DF, 
                  family = hurdle.negative.binomial(), zi_fixed = ~ sex)

hm1
```

Also for the `hurdle.negative.binomial()` family only the log-link is implemented. The 
structure of the output is identical to what we had for the hurdle Poisson model. 
Again we should note that the fixed-effects coefficients for the positive part have an 
interpretation for the mean of the entire distribution, not only for the positive counts.

We update again the fitted model `hm1` by including a random intercept for the zero-part:
```{r}
hm2 <- update(hm1, zi_random = ~ 1 | id)

hm2
```

The likelihood ratio test between the two models is computed with the `anova()` method:
```{r}
anova(hm1, hm2)
```

The result again indicates that the extra random effect for the zero-part improves the fit
of the model.

